Concurrent Programming in Harmony

### 2020-12-31 book/paper.tex
Latest

Some programming languages favor a different way of implementing
synchronization using so-called \emph{actors}~\cite{HBS73}.  Actors are
threads that have only private memory and communicate through
\emph{message passing}.
See \autoref{fig:actorpic} for an illustration.
Given that there is no shared memory in the actor model (other than the message
queues, which have built-in synchronization), there is no need
for critical sections.  Instead, some sequential thread owns a particular
piece of data and other threads access it by sending request messages
to the thread and optionally waiting for response messages.  Each thread
handles one message at a time, serializing all access to the data it owns.
As message queues are FCFS (First-Come-First-Served), starvation is prevented.

@@

Note that a \texttt{Queue} behaves much like a zero-initialized semaphore.
\texttt{put} is much like \texttt{V}, except that it is accompanied by data.
\texttt{get} is much like \texttt{P}, except that it also returns data.
Thus, synchronized queues can be considered a generalization of counting semaphores.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{figures/barrier-crop.pdf}
\end{center}
\caption{High-level state diagram specification of barrier synchronization
with three threads and two rounds.  The three numbers specify how many
rounds each thread has completed.}
\label{fig:barrierdiagram}
\end{figure}

@@

\autoref{fig:barrierdiagram} shows a high-level depiction of barrier synchronization
with three and threads.  Initially all threads are in round~0.  Then each thread
can start and finish the round.  However, none of the threads can
progress to the next round until all threads have finished the
current round.

@@

More generally, barrier synchronization can be abstracted as follows.
We want to create a \texttt{Barrier}$(n)$ object, with operations
\texttt{enter()} and \texttt{exit()}.  It is helpful to use
a roller coaster car with $n$ seats as a metaphor:
\begin{itemize}
\item the car cannot contain more than $n$ people;
\item the car won't take off until $n$ people are in the car;
\item no new people can enter the car until all $n$ people have left it.
\end{itemize}
Notice there are two different waiting conditions:
\begin{enumerate}
\item waiting for the car to empty out;
\item waiting for the car to fill up.
\end{enumerate}
Similarly, \texttt{enter}() blocks until $n$ threads have invoked it.
If more threads try to \texttt{enter}(), they will block until the
previous $n$ threads have invoked \texttt{exit}().

\autoref{fig:genbarrier} shows the code.
Method~\texttt{enter}() has two \texttt{wait} loops, one for each
waiting condition.
This is sometimes called \emph{double turnstile}.
Each loop uses a different condition variable.
The first loop waits while the barrier is fully occupied, while
the second waits for the barrier to fill up.
The $n^\mathit{th}$ thread to arrive at the barrier notifies the
$n-1$ threads that are waiting in the second loop.

Method \texttt{exit}() counts how many threads have left the
barrier.  When the last one leaves, it notifies all the threads
that are waiting in the first loop.

\section*{Exercises}
\begin{problems}
\item See if you can implement barrier synchronization for \texttt{N} threads
with just three binary semaphores.  Busy waiting is not allowed.  How about just two?
(As always, the Little Book of Semaphores~\cite{Downey09} is a good resource
for solving synchronization problems with semaphores.
Look for the \emph{double turnstile}
\index{double turnstile}%
solution.)
\end{problems}

@@

Operations on booleans include:
\begin{center}
\begin{tabular}{|l|l|}
\hline
\texttt{$e$ and $e$ and ...} & conjuction \\
\hline
\texttt{$e$ or $e$ or ...} & disjunction \\
\hline
\texttt{$e$ => $e$, $e$ not => $e$} & implication \\
\hline
\texttt{not $e$} & negation \\
\hline
\texttt{$v$ if $e$ else $v'$} & $v$ or $v'$ depending on $e$ \\
\hline
\texttt{any $s$, all $s$} & disjunction / conjunction for set $s$ \\
\hline
\end{tabular}
\end{center}

@@

@@

@@

You can denote a list by a sequence of values, each value terminated
by a comma.  As per usual, you can use brackets or parentheses
at your discretion.  For Python users, the important thing to note
is that a singleton lists in Harmony must contain a comma.  For
example $[1,]$ is a list containing the value $1$, while $[1]$
is simply the value $1$.

@@

@@

In Harmony, a string is a list of single-character atoms.
The Python string notations mostly work in Harmony as well.
For example, \texttt{'abc'} is a three-character string
consisting of the atoms \texttt{.a}, \texttt{.b}, and
\texttt{.c}.
\texttt{"abc'} is the same three-character string.
Multi-line strings can be terminated by triple quotes or
triple double-quotes.

@@

Harmony has native support for bags (aka multisets), which are sets
that can contain multiple copies of the same element.
It is represented by a dictionary that maps each element to its
multiplicity.
So \texttt{bag\{ 10, 10, 12 \} = dict\{ 10: 2, 12: 1 \}}.

@@

@@

A program counter is an integer that can be used to index into Harmony
bytecode.  When you define a method or a lambda function, you are
creating a constant of the program counter type.
Note, however, that labels are currently just integers.  This may
change in the future.

@@

\chapter{List of Operators}

Harmony currently supports the following operators:
\vspace{1em}

{\small
\begin{tabular}{|l|l|}
\hline
\texttt{$e$ == $e$}, \texttt{$e$ != $e$} & two Harmony values of the same type \\
\hline
\texttt{$e$ < $e$, $e$ <= $e$, $e$ > $e$, $e$ >= $e$} & any two Harmony values \\
\hline
\texttt{$e$ and $e$ and ...} & two or more booleans \\
\hline
\texttt{$e$ or $e$ or ...} & two or more booleans \\
\hline
\texttt{$e$ => $e$} & two booleans (implication) \\
\hline
\texttt{not $e$} & a boolean \\
\hline
\texttt{-$e$, \string~$e$, abs($e$)} & an integer \\
\hline
\texttt{$e$ + $e$ + ...} & two or more integers, tuples, or lists \\
\hline
\texttt{$e$ - $e$} & two integers or sets \\
\hline
\texttt{$e$ \& $e$ \& ..., $e$ | $e$ | ..., $e$ \string^ $e$ \string^ ...} & two or more integers or sets \\
\hline
\texttt{$e$ * $e$} & two integers or an integer and a list \\
\hline
\texttt{$e$ / $e$, $e$ // $e$} & two integers (integer division) \\
\hline
\texttt{$e$ \% $e$, $e$ mod $e$} & two integers (division remainder) \\
\hline
\texttt{$e$ ** $e$, $e << e$, $e >> e$} & two integers \\
\hline
\texttt{$\{ e..e \}$} & two integers \\
\hline
\texttt{$e$ [not] in $e$} & first is any Harmony value, second is a set or dict \\
\hline
\texttt{$e$ if $e$ else $e$} & middle $e$ must be a boolean \\
\hline
\texttt{!$e$} & an address \\
\hline
\texttt{choose $e$} & a set \\
\hline
\texttt{min $e$, max $e$, any $e$, all $e$, len $e$} & a set or a dict \\
\hline
\texttt{keys $e$} & a dictionary (includes tuple and list) \\
\hline
\texttt{atLabel $e$} & atom (corresponding to a label) \\
\hline
\texttt{get\_context $e$} & $e$ must be \texttt{()} \\
\hline
\texttt{setintlevel} \textit{e} & \textit{e} is a boolean \\
\hline
\texttt{stop} \textit{lv} & \textit{lv} is an lvalue \\
\hline
\texttt{?}\textit{lv} & \textit{lv} is an lvalue \\
\hline
\texttt{lambda $a$: $e$ end} & $a$ is a bounded variable, $e$ an expression\\
\hline
\end{tabular}
}
\vspace{1em}

\texttt{+} concatenates when applied to lists or tuples.
\texttt{-} computes set difference when applied to two sets.
\texttt{\&} computes the intersection when applied to sets.
\texttt{|} computes the union when applied to sets.
\texttt{\^} computes $(s \cup t) \backslash (s \cap t)$ when applied to sets $s$ and $t$.
\texttt{$\{ x .. y \}$} computes the set consisting of the integers $x$
through $y$ inclusive.  It returns the empty set if $x > y$.
If $d$ is a dictionary, $e$ \texttt{in} $d$ tests if $e$ is in the set of
\emph{values} of $d$, not the set of \emph{keys} (this is different from Python).

An \emph{lvalue}
\index{lvalue}%
(short for left hand value of an assignment statement)
is something that can be assigned.  This can be a
shared variable, a thread variable, or a dereferenced pointer
variable.  It can also be indexed.  Examples include
\textit{var}, \textit{var}[$e$], and \texttt{!}\textit{p}.

Harmony also supports the following comprehensions:

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
Set comprehension & \texttt{\{ f($v$) for $v$ in $s$ \}} \\
\hline
List comprehension & \texttt{[ f($v$) for $v$ in $s$ ]} \\
\hline
Dict comprehension & \texttt{dict\{ f($v$) for $v$ in $s$ \}} \\
\hline
\end{tabular}
\vspace{1em}

In each case, $s$ must be a set or a dictionary.
\autoref{ap:details} provides more information on comprehensions.

>>>>

Some programming languages favor a different way of implementing
synchronization using so-called \emph{actors}~\cite{HBS73}.  Actors are
threads that have only private memory and communicate through
\emph{message passing}.
See \autoref{fig:actorpic} for an illustration.
Given that there is no shared memory in the actor model (other than the message
queues, which have built-in synchronization), there is no need
for critical sections.  Instead, some sequential thread owns a particular
piece of data and other threads access it by sending request messages
to the thread and optionally waiting for response messages.  Each thread
handles one message at a time, serializing all access to the data it owns.
As message queues are FIFO (First-In-First-Out), starvation is prevented.

@@

For those familiar with counting semaphores: note that a \texttt{Queue}
behaves much like a zero-initialized counting semaphore.
\texttt{put} is much like \texttt{V}, except that it is accompanied by data.
\texttt{get} is much like \texttt{P}, except that it also returns data.
Thus, synchronized queues can be considered a generalization of
counting semaphores.

@@

@@

\begin{comment}
\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{figures/barrier-crop.pdf}
\end{center}
\caption{High-level state diagram specification of barrier synchronization
with three threads and two rounds.  The three numbers specify how many
rounds each thread has completed.}
\label{fig:barrierdiagram}
\end{figure}

\autoref{fig:barrierdiagram} shows a high-level depiction of barrier synchronization
with three and threads.  Initially all threads are in round~0.  Then each thread
can start and finish the round.  However, none of the threads can
progress to the next round until all threads have finished the
current round.
\end{comment}

@@

More generally, barrier synchronization can be abstracted as follows.
We want to create a \texttt{Barrier}$(n)$ object, with operations
\texttt{enter()} and \texttt{exit()}.  It is helpful to use
a roller coaster car with $n$ seats as a metaphor:
\begin{itemize}
\item the car cannot contain more than $n$ people;
\item the car won't take off until $n$ people are in the car;
\item no new people can enter the car until all $n$ people have left it.
\end{itemize}
Notice there are two different waiting conditions:
\begin{enumerate}
\item waiting for the car to empty out;
\item waiting for the car to fill up.
\end{enumerate}

\autoref{fig:genbarrier} shows the code.
Method~\texttt{enter}() has two \texttt{wait} loops, one for each
waiting condition.
This is sometimes called \emph{double turnstile}.
Each loop uses a different condition variable.
The first loop waits until the car has emptied out, while
the second waits for the car to fill up.

Notice how variables \textit{entered} and \textit{left}
are maintained.  They are incremented one-by-one, but they
are not decremented one-by-one.  Instead, the last thread to
enter sets \textit{left} to zero (and notifies the threads
that were waiting to for the car to fill up), while the last
thread to exit sets \textit{entered} to zero (and notifies
the threads that were waiting for the car to empty out).

\section*{Exercises}
\begin{problems}
\item Implement barrier synchronization for \texttt{N} threads
with just three binary semaphores.  Busy waiting is not allowed.
Can you implement barrier synchronization with two binary semaphores?
(As always, the Little Book of Semaphores~\cite{Downey09} is a good resource
for solving synchronization problems with semaphores.
Look for the \emph{double turnstile}
\index{double turnstile}%
solution.)
\item Imagine a pool hall with \texttt{N} tables.  A table is \emph{full}
from the time there are two players until both players have left.
When someone arrives, they can join a table that is not full, preferably
one that has a player ready to start playing.
Implement a simulation of such a pool hall.
\end{problems}

@@

Operations on booleans include:
\begin{center}
\begin{tabular}{|l|l|}
\hline
\texttt{$e$ and $e$ and ...} & conjuction \\
\hline
\texttt{$e$ or $e$ or ...} & disjunction \\
\hline
\texttt{$e$ => $e$, $e$ not => $e$} & implication \\
\hline
\texttt{not $e$} & negation \\
\hline
\texttt{$v$ if $e$ else $v'$} & $v$ or $v'$ depending on $e$ \\
\hline
\texttt{any $s$, all $s$} & disjunction / conjunction for set or list $s$ \\
\hline
\end{tabular}
\end{center}

@@

Harmony also supports \emph{set comprehension}.  In its simplest form,
\texttt{\{ f($v$) for $v$ in $s$ \}} returns a set that is constructed
by applying \texttt{f} to all elements in $s$ (where $s$ is a set or
a list).
This is known as \emph{mapping}.  But set comprehension is much more
powerful and can include joining multiple sets (using nested for
loops) and filtering (using the \texttt{such that} keywords).

For example:
\texttt{\{ $x$ + $y$ for $x$ in $s$ for $y$ in $s$ such that $(x * y) == 4$ \}}
returns a set that is constructed by summing pairs of elements from $s$
that, when multiplied, have the value 4.

@@

Dictionaries support comprehension.  The  basic form is:
\texttt{dict\{ f($v$) for $v$ in $s$ \}}.  This creates a dictionary
that maps each element $v$ in $s$ to $\mathtt{f}(v)$.

@@

You can denote a list by a sequence of values, each value terminated
by a comma.  As per usual, you can use brackets or parentheses
at your discretion.  For Python users, the important thing to note
is that a singleton list in Harmony must contain a comma.  For
example $[1,]$ is a list containing the value $1$, while $[1]$
is simply the value $1$.

@@

Lists and tuples support comprehension as well.
In their most basic form: \texttt{f($v$) for $v$ in $s$}.
For example, to check if any element in a list $t$ is even, you
can write:

\texttt{any($(x \% 2) == 0$ for $x$ in $t$)}.

@@

In Harmony, a string is a list of single-character atoms.
The Python string notations mostly work in Harmony as well.
For example, \texttt{'abc'} is a three-character string
consisting of the atoms \texttt{.a}, \texttt{.b}, and
\texttt{.c}.
\texttt{"abc"} is the same three-character string.
Various special characters (including quotes, newlines, etc.)
can be \emph{escaped} using a backslash.
Multi-line strings can be terminated by triple quotes or
triple double-quotes.

@@

Harmony has native support for bags (aka multisets), which are sets
that can contain multiple copies of the same element.
It is represented by a dictionary that maps each element to its
multiplicity.
So \texttt{bag\{ 10, 10, 12 \} = dict\{ 10:2, 12:1 \}}.

@@

Bags also support comprehension.

@@

A program counter is an integer that can be used to index into Harmony
bytecode.  When you define a method, a lambda function, or a label,
you are creating a constant of the program counter type.

Operations on program counters include the following:
\begin{center}
\begin{tabular}{|l|l|}
\hline
\texttt{$m$ $a$} & invoke method with program counter $m$ and argument $a$\\
\hline
\texttt{atLabel $l$} & return a bag of (method, argument) pairs of threads executing at label $l$ \\
\hline
\end{tabular}
\end{center}

You can create lambda functions similarly to Python.  For example:
\texttt{lambda$(x,y)$:$x+y$}.

@@

A thread can call \texttt{get\_context()} to retrieve its current
context.  Also, the function \texttt{contexts()} returns a
bag of all contexts (including terminated ones).
Both these functions should only be called in \texttt{atomic} mode
and used for checking invariants.
